---
title: "HRV analysis"
author: "Janice Tjeng"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

# Load packages
```{r}
library(RMySQL)
library(tidyverse)
library(lubridate)
library(robustbase)
library(stringr)
library(Rssa)
```

# Read data from E4, firstbeat, & MSband
```{r}
db <- dbConnect(MySQL(), user="deepresearcher", password="ctpc177!", dbname="deephealth2", host="deephealthlab.org")
e4 <- dbSendQuery(db,"SELECT * FROM viewl_e4_rr")
e4dat <- fetch(e4,n=-1) 
fb <- dbSendQuery(db, "SELECT * FROM viewl_firstbeat_rr")
fbdat <- fetch(fb, n=-1)
mb <- dbSendQuery(db, "SELECT * FROM viewl_msband_rr")
mbdat <- fetch(mb, n=-1)
```

# Read timing data
```{r}
timing <- read_csv("timing_data.csv")
df <- timing %>%
  select (Event, LWP2_0019, Activity) %>%
  rename("Time"="LWP2_0019")
n <- nrow(df)
dte <- mdy(as.character(df[1,"Time"]))
start <- as.character(df[2,"Time"])
end <- as.character(df[n,"Time"])
```

# Select rows for user 19 from E4 
```{r}
user19_l <- e4dat %>%
  filter(user=="lwp2_0019", device_location=="Left") %>% 
  mutate(readable_timestamp=as.POSIXct(readable_timestamp),
         e4_rr=e4_rr*1000) 
filter_time_day1_e4 <- user19_l %>%
  filter(strftime(readable_timestamp, "%H:%M:%S") >= start & strftime(readable_timestamp, "%H:%M:%S") <= end, date(readable_timestamp)==dte) 
```

# Select rows for user 19 from ECG
```{r}
user19_fb <- fbdat %>%
  filter(user=="lwp2_0019") %>% 
  mutate(readable_timestamp=as.POSIXct(readable_timestamp)) 
filter_time_day1_fb <- user19_fb %>%
  filter(strftime(readable_timestamp, "%H:%M:%S") >= start & strftime(readable_timestamp, "%H:%M:%S") <= end, date(readable_timestamp)==dte) 
```

# Select rows for user 19 from MSband 
```{r}
user19_l_mb <- mbdat %>%
  filter(user=="lwp2_0019", device_location=="Left") %>% 
  mutate(readable_timestamp=as.POSIXct(readable_timestamp),
         mb_rr=(mb_rr*1000)) 
filter_time_day1_mb <- user19_l_mb %>%
  filter(strftime(readable_timestamp, "%H:%M:%S") >= start & strftime(readable_timestamp, "%H:%M:%S") <= end, date(readable_timestamp)==dte) 
```

# Clean data

## Detect outlier, short RR-interval, based on z-score standardization. 

```{r}
# Standardize function, standard deviation based on median absolute deviation, a robust estimate of standard deviation
standardize <- function(rr){
  return ((rr-mean(rr))/sqrt(covMcd(rr)$cov[1])) # use robust covariance estimate
}

# Function to identify outlier
outlier_short <- function(rr){
  z <- standardize(rr)
  ind <- which(z<(-3)) # -3: limit in sd for outlier detection of short rr interval
  return (ind)
}

# Function to remove outlier, then add to subsequent beat
rm_short <- function(x,y){ # x is the index of outliers, y is the column of rr-interval
  for (i in 1:length(x)){
    if (x[i]<length(y)){
      y[x[i]+1] <- y[x[i]+1] + y[x[i]]
      y[x[i]] <- NA
    }
    else{  # if rr-interval is at the end of the array, just remove it without adding it to the next beat
      y[x[i]] <- NA
    }
  }
  return (y)
}

# Combine all functions
rm_short_comp <- function (rr){
  indx <- outlier_short(rr)
  clean <- rm_short(indx,rr)
  return (clean)
}
```

## Impute long rr-interval, applied after removing short
```{r}
impute_long <- function(rr){ 
  dmax <- mean(rr) + (2*sqrt(covMcd(rr)$cov[1]))
  dp <- numeric()
  dp1 <- numeric()
  ind_long <- which(rr>dmax) # Index of long outliers
  for (i in 1:length(ind_long)){     # imputation if rr>dmax
    dp <- sum((rr[ind_long][i]-dmax), dp)
  }
  dp <- dp/length(ind_long)          # imputation if rr<=dmax
  rr[ind_long] <- rr[ind_long]-dp
  for (i in 1:length(rr[-ind_long])){
    dp1 <- sum((rr[-ind_long][i]-dmax), dp1)
  }
  dp1 <- dp1/length(rr[-ind_long])
  rr[-ind_long] <- rr[-ind_long]-dp1
  return (rr)
}

```
https://www.r-bloggers.com/wheres-the-magic-emd-and-ssa-in-r/

https://www.researchgate.net/publication/228092069_Basic_Singular_Spectrum_Analysis_and_Forecasting_with_R

w-correlation matrix
**E4**
```{r}
filter_time_day1_e4.clean <- filter_time_day1_e4 %>%
  mutate(e4_rr=rm_short_comp(e4_rr)) %>%
  filter(!is.na(e4_rr)) %>%
  mutate(e4_rr=impute_long(e4_rr))
# 1st stage decomposition
e4.ssa <- ssa(filter_time_day1_e4.clean$e4_rr, kind="toeplitz-ssa", svd.method = "svd")
summary(e4.ssa)
# 1st stage visual information for grouping
plot(e4.ssa) # scree plot of singular values to identify trends and pairings. Trend at 1&4, 
# plot(e4.ssa, "series", groups=as.list(1:5)) # Reconstructed time series for first few components/projections
# 1st stage reconstruction
plot(e4.ssa, type="paired") #3&4, 6&7
res1 <- reconstruct(e4.ssa, groups=list(Tr=c(1,3,4,48)))
plot(res1, add.residuals=F) # Reconstructed series of the trend
res.trend <- residuals(res1) # Extract seasonality from residuals, original values-predicted values
spec.pgram(res.trend, detrend=F, log="no") # Periodogram, detect seasonality, should correspond to sine waves in paired plot

# Reconstructed plot
df_e4 <- data.frame(Reconstruct=res1$Tr, Original=filter_time_day1_e4.clean$e4_rr, Time=filter_time_day1_e4.clean$readable_timestamp)
ggplot(df_e4, aes(x=Time, y=Reconstruct)) + geom_line() + ylab("Reconstructed RR-interval in ms")

# Reconstructed and original plot combined
df_e4 %>%
  gather(Reconstruct, Original, key="Type", value="RR-interval") %>%
  ggplot(aes(x=Time, y=`RR-interval`)) +
  geom_line(aes(colour=Type)) +
  ylab("RR-interval in ms")

# Original plot
ggplot(filter_time_day1_e4.clean, aes(x=readable_timestamp, y=e4_rr)) + geom_line() +
  ggtitle("Plot of rr-interval against time, after removing outliers, \nmeasured by E4 on the left hand, \non first day of study, 2016-11- 15, for participant 19.") +
  ylab("RR interval in milliseconds") + 
  xlab("time")

```

**ECG**

```{r}
filter_time_day1_fb.clean <- filter_time_day1_fb %>%
  mutate(fb_rr=rm_short_comp(fb_rr)) %>%
  filter(!is.na(fb_rr)) %>%
  mutate(fb_rr=impute_long(fb_rr))
fb.ssa <- ssa(filter_time_day1_fb.clean$fb_rr,  kind="toeplitz-ssa", svd.method = "svd")
plot(fb.ssa) # Trends at 1 & 42
plot(fb.ssa, type="paired") #6&7, 7&8, 8&9, 9&10, 10&11
fb_res <- reconstruct(fb.ssa, group=list(Tr=c(1,42)))
plot(fb_res, add.residuals=F)

res.trend.fb <- residuals(fb_res) # Extract seasonality from residuals, original values-predicted values
spec.pgram(res.trend.fb, detrend=F, log="no") # Periodogram, detect seasonality, should correspond to sine waves in paired plot

# Reconstructed plot
df_fb <- data.frame(Reconstruct=fb_res$Tr, Original=filter_time_day1_fb.clean$fb_rr, Time=filter_time_day1_fb.clean$readable_timestamp)
ggplot(df_fb, aes(x=Time, y=Reconstruct)) + geom_line() + ylab("Reconstructed RR-interval in ms")

# Reconstructed and original plot combined
df_fb %>%
  gather(Reconstruct, Original, key="Type", value="RR-interval") %>%
  ggplot(aes(x=Time, y=`RR-interval`)) +
  geom_line(aes(colour=Type)) +
  ylab("RR-interval in ms")

# Original Plot
ggplot(filter_time_day1_fb.clean, aes(x=readable_timestamp, y=fb_rr)) + geom_line() +
  ggtitle("Plot of rr-interval against time, after removing outliers, \nmeasured by ECG during the first day of study, \n2016-11- 15, for participant 19.") +
  ylab("RR interval in milliseconds") + 
  xlab("time")
```

Huge spike of ECG at, 14:19:58, >37,000ms. This is due to removal of previous short rr-intervals (rows 4081-4177) and adding it to the rr-interval at row 4178. 

**MSband**

```{r}
filter_time_day1_mb.clean <- filter_time_day1_mb %>%
  mutate(mb_rr=rm_short_comp(mb_rr)) %>%
  filter(!is.na(mb_rr)) %>%
  mutate(mb_rr=impute_long(mb_rr))

mb.ssa <- ssa(filter_time_day1_mb.clean$mb_rr,  kind="toeplitz-ssa", svd.method = "svd")
plot(mb.ssa) # Trends at 1 & 3
plot(mb.ssa, type="paired") # 3&4
mb_res <- reconstruct(mb.ssa, group=list(Tr=c(1,3,46)))
plot(mb_res, add.residuals=F)

res.trend.mb <- residuals(mb_res) # Extract seasonality from residuals, original values-predicted values
spec.pgram(res.trend.mb, detrend=F, log="no") # Periodogram, detect seasonality, should correspond to sine waves in paired plot

# Reconstructed plot
df_mb <- data.frame(Reconstruct=mb_res$Tr, Original=filter_time_day1_mb.clean$mb_rr, Time=filter_time_day1_mb.clean$readable_timestamp)
ggplot(df_mb, aes(x=Time, y=Reconstruct)) + geom_line() + ylab("Reconstructed RR-interval in ms")

# Reconstructed and original plot combined
df_mb %>%
  gather(Reconstruct, Original, key="Type", value="RR-interval") %>%
  ggplot(aes(x=Time, y=`RR-interval`)) +
  geom_line(aes(colour=Type)) +
  ylab("RR-interval in ms")


# Original Plot
ggplot(filter_time_day1_mb.clean, aes(x=readable_timestamp, y=mb_rr)) + geom_line() +
  ggtitle("Plot of rr-interval against time, after removing outliers, \nmeasured by MSband on the left hand, \non first day of study, 2016-11- 15, for participant 19.") +
  ylab("RR interval in milliseconds") + 
  xlab("time")
```

# RMSSD equation
```{r}
# x should be the column containing rr-interval, units in millisecond
HRV <- function(x){
  total <- 0
  for (i in 1:(length(x)-1)){
     total <- sum(total, (x[i+1] -x[i])^2, na.rm=T)
  }
  return (sqrt(total/(length(x)-1)))
}
```

# Select each period and calculate RMSSD for E4, fb, and MSband
```{r}
for (i in 3:(n-1)){
  # select time for rr-interval 
  rr_int_e4 <- df_e4 %>%
    filter(strftime(Time, "%H:%M")>=df$Time[i] & strftime(Time, "%H:%M") <df$Time[i+1])
  rr_int_fb <- df_fb %>%
    filter(strftime(Time, "%H:%M")>=df$Time[i] & strftime(Time, "%H:%M") < df$Time[i+1])
  rr_int_mb <- df_mb %>%
    filter(strftime(Time, "%H:%M")>=df$Time[i] & strftime(Time, "%H:%M") < df$Time[i+1])
  # calculate RMSSD for each rr interval according to time
  df$RMSSD_e4[i] <- HRV(rr_int_e4$Reconstruct)
  df$RMSSD_fb[i] <- HRV(rr_int_fb$Reconstruct)
  df$RMSSD_mb[i] <- HRV(rr_int_mb$Reconstruct)
}
df$RMSSD_e4[n] <- NA
df$RMSSD_fb[n] <- NA
df$RMSSD_mb[n] <- NA

df #output
```

# Plot HRV over different periods for each device

```{r}
hrv_df <- df %>%
  slice(3:20) %>%
  gather(`RMSSD_e4`, `RMSSD_fb`, `RMSSD_mb`, key="Device", value="RMSSD") %>%
  mutate(Device=str_replace(Device, "^.*_", ""),
         Event=factor(Event, levels=df$Event[3:20]))

clr <- c("blue", "green", "purple", "red", "black", "purple", "blue", "green", "purple", "orange", "blue", "gray", "purple", "yellow", "pink", "black", "purple", "violet")
ggplot(hrv_df, aes(x=Event, y=RMSSD)) +
  geom_path(aes(group=Device)) +
  geom_point(aes(colour=Event)) +
  facet_wrap(~Device) +
  scale_colour_manual(values=clr, 
                      labels=paste(hrv_df$Event,hrv_df$Activity, sep=":"),
                      name=NULL) +
  #annotate(geom="text", x=1, y=50, label="E4") +
  #annotate(geom="text", x=1, y=30, label="ECG") +
  #annotate(geom="text", x=1, y=100, label="MSB") +
  ggtitle("Plot of RMSSD across events for each device") +
  ylab("RMSSD in milliseconds") +
  theme(legend.position="top", 
        legend.text=element_text(size=7),
        axis.text.x = element_text(angle = 90, hjust = 1)) 
```

# Implement timing, 3 min + 30s 
```{r}
time <- hms(start) + minutes(3)
timings <- c(as.character(hms(start)),vector())
while (as.character(time)<(hms(end)-seconds(30))){
  timings <- c(timings, as.character(seconds_to_period(period_to_seconds(time)+seconds(30))))
  time <- seconds_to_period(period_to_seconds(time)+seconds(30))
}
timings <- c(timings, as.character(hms(end)))
for (i in 1:length(timings)){
  timings[i]<- ifelse(nchar(timings)[i]<11,
                      paste(substr(timings[i],1,3)," 0", substr(timings[i],5,nchar(timings[i])), sep=""),
                      timings[i])
  # Add zeros in between minutes for some timings
}
timings <- substr(str_replace_all(timings, "\\D\\s", ":"),1,8) # Replace HMS with semicolon for comparison of timing
timings
df_window <- data.frame(time=timings) %>%
  mutate(time=as.character(time))
for (i in 1:(nrow(df_window)-1)){
  # select time
  e4 <- df_e4 %>%
    filter(strftime(Time, "%H:%M:%S")>=df_window$time[i] & strftime(Time, "%H:%M:%S") <df_window$time[i+1])
  fb <- df_fb %>%
  filter(strftime(Time, "%H:%M:%S")>=df_window$time[i] & strftime(Time, "%H:%M:%S") < df_window$time[i+1])
  mb <- df_mb %>%
  filter(strftime(Time, "%H:%M:%S")>=df_window$time[i] & strftime(Time, "%H:%M:%S") < df_window$time[i+1])
  # calculate RMSSD for each time interval
  df_window$RMSSD_e4[i] <- HRV(e4$Reconstruct)
  df_window$RMSSD_fb[i] <- HRV(fb$Reconstruct)
  df_window$RMSSD_mb[i] <- HRV(mb$Reconstruct)
}

df_window$RMSSD_e4[146] <- 0
df_window$RMSSD_e4[df_window$RMSSD_e4==0] <- NA
head(df_window)
tail(df_window)
```

Zeros or NaN in between indicate that there isn't any value for certain timings

# Plot HRV across timing
```{r}
df_window1 <- df_window %>%
   gather(`RMSSD_e4`, `RMSSD_fb`, `RMSSD_mb`, key="Device", value="RMSSD") %>%
   mutate(Device=str_replace(Device, "^.*_", ""))
head(df_window1)

ggplot(df_window1, aes(x=time,y=RMSSD)) +
  geom_path(aes(group=Device)) +
  facet_wrap(~Device) 

```
e4 disconnected due to NAs.